Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-23 paper.tex
todo
untabify

\subsection{Coercions in Normal Form} 
\label{sec:coercion-calculus}

In this section we review the
coercions~\citep{henglein1994dynamic,herman2010space} and the normal
form of \citet{siek2012interpretations} to motivate the design of
hypercoercions.  We ignore sum types and product types in this
section, because \citet{siek2012interpretations} did not discuss
them. We assume a basic familiarity with coercions and direct readers
who are not familiar with them to \citet{siek2012interpretations}.

@@

\figref{fig:normal-coercion} reviews the grammar for coercions,
written $c$.
%
To review, an injection $I!$ takes a value from an injectiable type
$I$ to type $\TOOdyn$. An injectable type is simply a type that can be
cast directly to (injection) and from (projection) $\TOOdyn$. The
definition of injectable type depends on blame strategy. With \lazyD,
all pre-types are injectable. With \lazyUD, only
$\POOfun{\TOOdyn}{\TOOdyn}$ is injectable.
%
A projection $I?^l$ takes a value from type $\TOOdyn$ to type $I$, or
halts the program and blames $l$ if the value is of a different type.
The coercion $\iota$ is the identity, $\bot^{l}$ is the coercion that
always fails and blames $l$, and $\ncSeq{c_1}{c_2}$ applies $c_1$ and
then $c_2$ in sequence. The function coercion $\ncFun{c_1}{c_2}$
applies $c_1$ to the argument of a function and $c_2$ to the return
pvalue.

Coercions come with a reduction relation so it is natural to ask what
are their normal forms. \figref{fig:normal-coercion} defines the
syntax of coercion in normal form, written $\hat{c}$. Conceptually, a
coercion in normal form has three parts, all of which are optional.
It may start with a projection $I?^l$, then it may be followed by a
function coercion, and then it may be concluded with an injection $I!$
or a failure $\bot^l$. While this is a simple idea, there are 10
clauses in the definition for $\hat{c}$!

@@

	hypercoercion typing \fbox{$c : T \Longrightarrow T$}
	\begin{gather*}
	\inference{}{\typingHC{\hyperCoercionI}{\TOOdyn}{\TOOdyn}}
	\quad
	\inference{
		\typingHC{h}{T_1}{P_1} &
		\typingHC{m}{P_1}{P_2} &
		\typingHC{t}{P_2}{T_2}
	}{
		\typingHC{\hyperCoercionC{h}{m}{t}}{T_1}{T_2}
	}
	\end{gather*}

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observation that a normal coercion has at most three parts;
projections are always at the beginning; injections and failures are
always at the end. Hypercoercions make this structure explicit: a
hypercion either is $\hyperCoercionI$, the identity cast for
$\TOOdyn$, or contains three parts: the head is either a projection or
the identity, the tail is either an injection, a failure, or the
identity, and the middle involves coercions underneath a type
constructor, i.e., coercions between function, pair, and sum types.

\figref{fig:hypercoercion} defines the syntax of
hypercoercions. Types, pre-types, and blame labels are as before.  Let
$c$ range over hypercoercions. A hypercoercion either is the identity
cast between the dynamic types, or includes a head, a middle, and a
tail.  Let $h$ range over heads. A head is either a no-op, or a
projection.  Let $m$ range over middles. There is a one-to-one
correspondence between middles and type constructors.  We generalize
shallow-consistency to middles $m \smile m$ in the obvious way.  Let
$t$ range over tails. A tail is either a no-op, an injection, or a
failure.

Subsection~\ref{sec:ld-hc} defines functions that construct \lazyD\ 
hypercoercions. Subsection~\ref{sec:lud-hc} defines the \lazyUD\ counterparts. 
Functions that apply hypercoercions to values are deferred to 
Section~\ref{sec:hypercoercion-correctness} because they depends on a new 
definition of values, which is part of the framework in
Section~\ref{sec:framework}.

@@

>>>>

\subsection{Coercions and Normal Forms} 
\label{sec:coercion-calculus}

In this section we review the
coercions~\citep{henglein1994dynamic,herman2010space} and the normal
form of \citet{siek2012interpretations} to motivate the design of
hypercoercions.  We omit sum types and product types in this section,
because \citet{siek2012interpretations} did not discuss them. We
assume a basic familiarity with coercions and direct readers who are
not familiar with them to \citet{siek2012interpretations} and
\citet{siek2015blame}.

@@

\figref{fig:normal-coercion} reviews the grammar for coercions,
written $c$.
%
To review, an injection $I!$ takes a value from an injectiable type
$I$ to type $\TOOdyn$. An injectable type is simply a type that can be
cast directly to (injection) and from (projection) $\TOOdyn$. The
definition of injectable type depends on blame strategy. With \lazyD,
all pre-types are injectable. With \lazyUD, only
$\POOfun{\TOOdyn}{\TOOdyn}$ is injectable.
%
A projection $I?^l$ takes a value from type $\TOOdyn$ to type $I$, or
halts the program and blames $l$ if the value is of a different type.
The coercion $\iota$ is the identity, $\bot^{l}$ is the coercion that
always fails and blames $l$, and $(\ncSeq{c_1}{c_2})$ applies $c_1$ and
then $c_2$ in sequence. The function coercion $\ncFun{c_1}{c_2}$
applies $c_1$ to the argument of a function and $c_2$ to the return
pvalue.

Coercions come with a reduction relation so it is natural to ask about
their normal forms. \figref{fig:normal-coercion} defines the syntax of
coercion in normal form, written $\hat{c}$. Conceptually, a coercion
in normal form has three parts, all of which are optional.  It may
start with a projection $I?^l$, followed by a function coercion, and
then concluded with an injection $I!$ or a failure $\bot^l$. While
this is a simple idea, there are 10 clauses in the definition for
$\hat{c}$!

@@

  Hypercoercion typing \fbox{$c : T \Longrightarrow T$}
  \begin{gather*}
  \inference{}{\typingHC{\hyperCoercionI}{\TOOdyn}{\TOOdyn}}
  \quad
  \inference{
    \typingHC{h}{T_1}{P_1} &
    \typingHC{m}{P_1}{P_2} &
    \typingHC{t}{P_2}{T_2}
  }{
    \typingHC{\hyperCoercionC{h}{m}{t}}{T_1}{T_2}
  }
  \end{gather*}

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observation that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the identity,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the identity.
\end{itemize}
There is a one-to-one correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.

Subsection~\ref{sec:ld-hc} defines functions that construct
\lazyD\ hypercoercions. Subsection~\ref{sec:lud-hc} defines the
\lazyUD\ counterparts.  The functions that apply hypercoercions to
values are defined later in
Section~\ref{sec:hypercoercion-correctness}, after we introduce the
space-efficient machine and its definition of values in
Section~\ref{sec:framework}.

@@

\todoKC{write conclusion. -KC}
